Geospatial - Assignment 2

### 2025-02-04 hw2/hw2_MoritzPeistNoemiLucchiSimonVellin.Rmd
Fixed markdown

```{r}
# Remove Seven seas and Antarctica because they are NA
sf.world <- sf.world %>%
  filter(!is.na(continent)) %>%
  # Repair invalid polygons
  st_make_valid() %>%
  # Create the population ranges as factors
  mutate(pop_range = factor(
    case_when(
      POP_EST < 1e7 ~ "0-10M",
      POP_EST >= 1e7 & POP_EST < 5e7 ~ "10M-50M",
      POP_EST >= 5e7 & POP_EST < 1e8 ~ "50M-100M",
      POP_EST >= 1e8 & POP_EST < 5e8 ~ "100M-500M",
      TRUE ~ "500M+"
    ),
    levels = c("0-10M", "10M-50M", "50M-100M", "100M-500M", "500M+")
  ))

@@

```{r}
sf.world_filt <- sf.world %>%
  mutate(
    continent = case_when(
      continent %in% c("Seven seas (open ocean)", "Antarctica") ~ NA_character_,
      TRUE ~ continent
    )
  ) %>%
  filter(!is.na(continent))

@@

```{r}
distances_filt <- distances %>%
  mutate(
    continent = case_when(
      continent %in% c("Seven seas (open ocean)", "Antarctica") ~ NA_character_,
      TRUE ~ continent
    )
  ) %>%
  filter(!is.na(continent))

@@

```{r}
# 3. Calculate average distances at country level. With this approach, instead of
# computing the closest airport distance per city and then averaging per country,
# we compute all the distances within a country and then we average them.
distances_country <- cities_with_countries %>%
  group_by(continent, iso_a2) %>% # Removed ne_id to group at country level
  summarise(
    avg_closest_distance_km = {
      current_iso <- cur_group()$iso_a2
      country_airports <- airports_with_countries %>%
        filter(iso_a2 == current_iso)
      if (nrow(country_airports) == 0) {
        NA_real_
      } else {
        closest_distances <- st_distance(
          geometry,
          country_airports
        ) %>%
          apply(1, min) %>%
          as.numeric() / 1000
        mean(closest_distances, na.rm = TRUE)
      }
    },
    country_name = first(name)
  ) %>%
  ungroup() %>%
  drop_na()

@@

The analysis is structured in three main parts:
1. Visualization of market locations across Africa
2. Analysis of markets' accessibility through distance calculations
3. Investigation of price-distance relationships for different transportation modes

@@

This section creates a visualization of market locations across Africa. The analysis combines three key geographical datasets:
- Administrative boundaries of African countries
- Road network data for transportation infrastructure
- Market location coordinates from the Porteous dataset

@@

The plotting phase creates a clear visualization of market locations within their geographical context. We use a layered approach to build the map, starting with country boundaries and adding market locations. The aesthetic choices are deliberate:
- Antiquewhite fill for countries provides a neutral background
- Gray borders help define country boundaries without overshadowing the main data
- Red and blue markers for markets ensure visibility while distinguishing them from the background
- Minimal theme removes unnecessary visual elements to focus attention on the data

@@

This subsection expands our dataset by incorporating additional infrastructure information. We load two new spatial datasets:
1. Airport locations across the globe
2. Coastline geometries

@@

This crucial data processing step involves several sophisticated spatial operations. The process:
1. Isolates African data to improve computational efficiency
2. Ensures geometric validity of spatial features
3. Calculates distance matrices between markets and various infrastructure points
4. Processes these matrices to find minimum distances

@@

The resulting visualizations help us understand whether more isolated markets (those farther from transportation infrastructure) tend to have higher or lower prices. This relationship is crucial for:
- Understanding market integration
- Identifying potential barriers to trade
- Assessing the impact of infrastructure on market efficiency

@@

This phase processes the raw price data to create meaningful economic metrics. The operations include:
1. Handling zero values by converting them to NA to avoid skewing averages
2. Calculating average prices across different time periods
3. Joining price data with spatial market data

@@

The final visualization phase creates three sophisticated scatter plots that reveal relationships between prices and infrastructure accessibility. Each plot incorporates:
1. Individual data points colored by crop type to show product-specific patterns
2. Trend lines to highlight overall relationships
3. Appropriate axis labels and titles for clear interpretation

>>>>

```{r}
# Filter countries with NA continents and repair invalid polygons
sf.world <- sf.world %>%
  filter(!is.na(continent)) %>%
  # Repair invalid polygons
  st_make_valid() %>%
  # Create the population ranges as factors
  mutate(pop_range = factor(
    case_when(
      POP_EST < 1e7 ~ "0-10M",
      POP_EST >= 1e7 & POP_EST < 5e7 ~ "10M-50M",
      POP_EST >= 5e7 & POP_EST < 1e8 ~ "50M-100M",
      POP_EST >= 1e8 & POP_EST < 5e8 ~ "100M-500M",
      TRUE ~ "500M+"
    ),
    levels = c("0-10M", "10M-50M", "50M-100M", "100M-500M", "500M+")
  ))

@@

```{r}
# Filter out Seven Seas and Antarctica
sf.world_filt <- sf.world %>%
  mutate(
    continent = case_when(
      continent %in% c("Seven seas (open ocean)", "Antarctica") ~ NA_character_,
      TRUE ~ continent
    )
  ) %>%
  filter(!is.na(continent))

@@

```{r}
# Filter out Seven Seas and Antarctica
distances_filt <- distances %>%
  mutate(
    continent = case_when(
      continent %in% c("Seven seas (open ocean)", "Antarctica") ~ NA_character_,
      TRUE ~ continent
    )
  ) %>%
  filter(!is.na(continent))

@@

```{r}
# 3. Calculate average distances at country level. With this approach, instead of
# computing the closest airport distance per city and then averaging per country,
# we compute all the distances within a country and then we average them.

# Basically, same code as before, so no extensive comments
distances_country <- cities_with_countries %>%
  group_by(continent, iso_a2) %>% # Removed ne_id to group at country level
  summarise(
    avg_closest_distance_km = {
      current_iso <- cur_group()$iso_a2
      country_airports <- airports_with_countries %>%
        filter(iso_a2 == current_iso)
      if (nrow(country_airports) == 0) {
        NA_real_
      } else {
        closest_distances <- st_distance(
          geometry,
          country_airports
        ) %>%
          apply(1, min) %>%
          as.numeric() / 1000
        mean(closest_distances, na.rm = TRUE)
      }
    },
    country_name = first(name)
  ) %>%
  ungroup() %>%
  drop_na()

@@

The analysis is structured in three main parts:

1. Visualization of market locations across Africa
2. Analysis of markets' accessibility through distance calculations
3. Investigation of price-distance relationships for different transportation modes

@@

This section creates a visualization of market locations across Africa. The analysis combines three key geographical datasets:

- Administrative boundaries of African countries
- Road network data for transportation infrastructure
- Market location coordinates from the Porteous dataset

@@

The plotting phase creates a clear visualization of market locations within their geographical context. We use a layered approach to build the map, starting with country boundaries and adding market locations. The aesthetic choices are deliberate:

- Antiquewhite fill for countries provides a neutral background
- Gray borders help define country boundaries without overshadowing the main data
- Red and blue markers for markets ensure visibility while distinguishing them from the background
- Minimal theme removes unnecessary visual elements to focus attention on the data

@@

This subsection expands our dataset by incorporating additional infrastructure information. We load two new spatial datasets:

1. Airport locations across the globe
2. Coastline geometries

@@

This crucial data processing step involves several sophisticated spatial operations. The process:

1. Isolates African data to improve computational efficiency
2. Ensures geometric validity of spatial features
3. Calculates distance matrices between markets and various infrastructure points
4. Processes these matrices to find minimum distances

@@

The resulting visualizations help us understand whether more isolated markets (those farther from transportation infrastructure) tend to have higher or lower prices. This relationship is crucial for:

- Understanding market integration
- Identifying potential barriers to trade
- Assessing the impact of infrastructure on market efficiency

@@

This phase processes the raw price data to create meaningful economic metrics. The operations include:

1. Handling zero values by converting them to NA to avoid skewing averages
2. Calculating average prices across different time periods
3. Joining price data with spatial market data

@@

The final visualization phase creates three sophisticated scatter plots that reveal relationships between prices and infrastructure accessibility. Each plot incorporates:

1. Individual data points colored by crop type to show product-specific patterns
2. Trend lines to highlight overall relationships
3. Appropriate axis labels and titles for clear interpretation
